Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization

### 2013-05-10 paper.tex
Added short overviews of what PDCFA and CFA2 actually do before going into our own presentation.

\usepackage{amsmath,stmaryrd,natbib,listings}
\usepackage[usenames,dvipsnames]{color}

@@

Testing new ideas in analysis for improving precision or increasing
performance can be a difficult venture. We contend that the machinery
that we employ in the abstract should have a concrete counterpart that
maintains the meaning of the language so that we can see their effect
without introducing approximations which might mask correctness
issues. Abstraction should be a simple process that is ``obviously
correct.'' The framework we present in this paper is applicable in the
concrete such that a pointwise abstraction leads to the pushdown
analyses in the literature.

@@

There are common underpinnings of PDCFA and CFA2 that can be embodied
as concrete machinery in the programming language semantics: 
\begin{enumerate}
\item{breaking the recursive structure of continuations by indirecting them through a
table with appropriately precise keys, and}
\item{memoization.}
\end{enumerate}
Once the machine semantics is in this form, simple pointwise
abstraction leads to the summarization algorithms that we see in the
literature.

@@

PDCFA does not have some orthogonal semantic components that CFA2
features to improve precision and is thus the simpler of the two. We
start by deriving PDCFA from an operational semantics for the untyped
lambda calculus (figure \ref{fig:base-semantics}). The semantic spaces
for the machine follow.

@@

The $\Addr$ space is what controls the precision of the model. For a
concrete semantics, we require the allocation metafunction $\alloc :
\State \to \Addr$ to return fresh addresses ($\maddr \notin
\dom(\mstore)$), but any choice of address is sound. This metafunction
along with the commitment to having no recursive datastructures is the
key to the AAM technique. All recursion can be expressed by
indirecting through addresses into a store --- tying Landin's knot. We
borrow this technique on top of our own.

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont}$
    &
    $\tpl{\mval,\mstore,\mkont}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kar{\mexpri,\menv,\mkont}}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kar{\mexpr,\menv,\mkont}}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kfn{\mval, \mkont}}$
    \\
% function call
    $\tpl{\mval,\mstore,\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}, \mkont}}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\mval}, \mkont}$
  \end{tabular}
  \caption{The CESK machine}
  \label{fig:base-semantics}
\end{figure}

Since PDCFA is a simpler analysis than CFA2, it enjoys a polynomial
complexity in the monovariant case. Its final ${\mathcal O}(n^6)$ form
is a naive ``step all states'' widening, which yields poor
results. The common widening uses a worklist to only step states that
need stepping --- the fact that the store is shared amongst all states
makes this a semantic difference, and not just an implementation
difference. Doing this bumps the complexity up to ${\mathcal
  O}(n^9)$. The available \emph{implementations} of PDCFA indeed do
this and are themselves ${\mathcal O}(n^9)$. These bounds are (rough)
absolute worst case that is highly unlikely, if ever, happen in
practice. The worklist abstraction restricts value flow to fewer
spurious paths, often leading to a faster analysis than a typical
${\mathcal O}(n^3)$ 0CFA.

\paragraph{Tracking return points:} the magic of the method is in
keeping a table of (local) continuations for each function and store
pair. This closely mirrors the technique of store-allocating
continuations used in AAM. Instead of deferring to the table (store)
for the remainder of a continuation for each frame, we only have
indirections at function call boundaries (see rules 4 and 5 in figure
\ref{fig:summary-semantics} for saving and restoring
continuations). Note that since AAM is about finitizing the state
space, this step itself fits within the AAM method, since
continuations within functions truncated at call boundaries are
bounded by the nesting depth of those functions, thereby making the
continuation space finite. In the monovariant case, the number of
continuations is still linear in the size of the program.

The tail of a continuation in the context of a function will contain
the stackless context in which the function was called, in order to
link up with the proper call-site(s). The context includes the
function (or unique label of the function), the environment, and the
store. Notice that this choice can be viewed as a special allocation
strategy for continuations in the AAM viewpoint; the store is the
important ingredient to maintaining enough precision to keep a
pushdown abstraction. We separate the concepts in this paper for the
exposition. The semantic spaces for the machine are modified thusly:

@@

@@

CFA2 is the first published analysis of a higher-order programming
language that could properly match calls and returns. We will show
that it fits well into the same presentation we gave for
PDCFA. Vardoulakis and Shivers had a clear goal of harnessing the
extra information a pushdown model provides to produce a
high-precision analysis that works well in practice. This resulted in
more than just the call/return matching of the previous section, which
is why we are showing the two separately. There are two orthogonal
features of the semantics:
\begin{enumerate}
\item{stack allocation for some bindings}
\item{strong updates on stack frames for resolved nondeterminism}
\end{enumerate}
The first of these is an addition to the stackless context. There is a
conservative pre-analysis that checks locally whether a binding will
never escape, and classifies references as able to use the stack frame
or not. Their criteria for a binding never escaping is that it is
never referenced in a function that is not its binder. This can be
extended in a language with more linguistic features; see Kranz's
thesis about register-allocatable bindings in the Orbit Scheme
compiler \citep{ianjohnson:kranz:thesis:1988}. A stackable reference
is one that appears in the body of the binding function, by which we
mean not within the body of a nested function. They use the information
that a binding never escapes to not bother allocating it in the
heap. This has the advantage of not changing the heap, and thus leads
to less propagation. The addition of these stack frames makes the
analysis exponential in theory, though in practice they have been
observed to decrease running time in most cases.

@@

@@

The metafunctions the semantics uses to entend and consult the heap
and stack use implicit information from the preanalysis we described
above:

@@

Sidestepping fake rebinding does not need to be restricted to
stackable references, but that is what CFA2 does. Indeed, as soon as
the nondeterminism has been determined, we could extend the stack
frame so any subsequent reference means what it meant previously in
the function. Lookup would then always try the stack frame before
falling back on the heap.

@@

The metafunctions mentioned in the fixed semantics all deal with the
addition of $\maddr$ to contexts. If the context in an $\rt$ frame is
approximate, we must return to all the continuations known for all the
contexts it approximates ($\returns$). At return boundaries, the memo
table must add the result to all the represented contexts
($\memoize$). At capture time, we strip the store in the $\rt$ frame
if there is one, and replace it with an address.

@@

A new temporal logic that itself understands well-bracketing of call
and returns for stating and validating propositions, NWTL
~\citep{ianjohnson:alur:nwtl:2007}, applies to visibly pushdown
systems, which can precisely model progamming languages that only have
well-bracketed call and returns. The propositional form of the logic
has decidable satisfiability checking, and control-flow queries are
shallow propositions to pose in the logic. However, the satisfiability
problem is \textsc{2Exptime}-hard and the algorithm is not given in
the on-the-fly form that a higher-order language would need to
effectively answer control-flow queries without already knowing the
answers. The logic itself is an exciting new frontier for expressing
program correctness properties, and a higher-order version would be a
welcome addition to the functional programmer's toolbelt.

@@

There is also the question of clients of these analyses. Most
obviously we would want to know where we can implement first-class
control more efficiently. In particular, we would want to have an
escape analysis to find first-class continuations that don't need to
be heap-allocated, and single target abort operations that can be
turned into long jumps after a computed stack unraveling. On top of
optimizations, there are security analyses. Greater control of the
stack gives us the ability to drill deep into context-sensitive
security properties and make precise predictions such as, ``is it
possible to send image data on the network outside the context of the
user's permission?'' This way a user can use an image taking/sharing
application without the application surrepticiously taking and sending
images the user doesn't know about.

>>>>

\pagestyle{plain}
\usepackage{amsmath,stmaryrd,natbib,listings}
\usepackage[usenames,dvipsnames]{color}

@@

Testing new ideas in analysis for improving precision or increasing
performance can be a difficult venture. We contend that the machinery
that we employ in the abstract should have a concrete counterpart that
maintains the meaning of the language so that we can see their effect
without introducing approximations which might mask correctness
issues. Abstraction should be a simple process that is ``obviously
correct.'' The framework we present in this paper is applicable in the
concrete such that a point-wise abstraction leads to the pushdown
analyses in the literature.

@@

There are common underpinnings of PDCFA and CFA2 that can be embodied
as concrete machinery in the programming language semantics: 
\begin{enumerate}
\item{breaking the recursive structure of continuations by indirecting them through a
table with appropriately precise keys, and}
\item{memoization.}
\end{enumerate}
Once the machine semantics is in this form, simple point-wise
abstraction leads to the summarization algorithms that we see in the
literature.

@@

PDCFA does not have some orthogonal semantic components that CFA2
features to improve precision and is thus the simpler of the two. The
key to their method is to notice in languages without stack-capturing
features, the continuation is only modified a frame at a time. If the
state space without the stack is finite (and it can be made finite
with an approximation ala AAM), then the model falls directly into the
realm of pushdown systems. They thus recast the problem in terms of
pushdown systems. The whole of their machinery is then computing the
pushdown system on-the-fly, only considering states reachable from the
root (initial) state. To match nodes that push frames with later nodes
following the pop of that frame, they additionally compute an
$\epsilon$-closure graph, which keeps edges between nodes that have a
net-zero stack change path between them. We skip the detour to
pushdown systems altogether and show an operational understanding of
the resulting analysis.

We start by deriving PDCFA from an operational semantics for the
untyped lambda calculus (figure \ref{fig:base-semantics}). The
semantic spaces for the machine follow.

@@

The $\Addr$ space is what controls the precision of the model. For a
concrete semantics, we require the allocation meta-function $\alloc :
\State \to \Addr$ to return fresh addresses ($\maddr \notin
\dom(\mstore)$), but any choice of address is sound. If $\alloc$ only
uses addresses from a finite subset of $\Addr$, then the state space
without continuations is finite, and the space of contination frames
is finite (thus the pushdown system interpretation is apt). This
meta-function along with the commitment to having no recursive
data-structures is the key to the AAM technique. All recursion can be
expressed by indirecting through addresses into a store --- tying
Landin's knot. We borrow this technique on top of our own.

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont}$
    &
    $\tpl{\mval,\mstore,\mkont}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kar{\mexpri1,\menv,\mkont}}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kar{\mexpr,\menv,\mkont}}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kfn{\mval, \mkont}}$
    \\
% function call
    $\tpl{\mval,\mstore,\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}, \mkont}}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\mval}, \mkont}$
  \end{tabular}
  \caption{The CESK machine}
  \label{fig:base-semantics}
\end{figure}

\paragraph{Tracking return points:} the magic of the method is in
keeping a table of continuation segments for each function and store
pair, where the continuation is segmented at function boundaries. This
closely mirrors the technique of store-allocating continuations used
in AAM. Instead of deferring to the table (store) for the remainder of
a continuation for each frame, we only have indirections at function
call boundaries (see rules 4 and 5 in figure
\ref{fig:summary-semantics} for saving and restoring
continuations). Note that since AAM is about finitizing the state
space, this step itself fits within the AAM method, since
continuations within functions truncated at call boundaries are
bounded by the nesting depth of those functions, thereby making the
continuation space finite. In the monovariant case, the number of
continuations is still linear in the size of the program.

The tail of a continuation in the context of a function will contain
the stack-less context in which the function was called, in order to
link up with the proper call-site(s). The context includes the
function (or unique label of the function), the environment, and the
store. Notice that this choice can be viewed as a special allocation
strategy for continuations in the AAM viewpoint, but we separate the
table of continuations from the store since continuations themselves
contain stores in $\rt$ frames --- this leads to a recursive
data-structure that we are trying to avoid. The store is an important
ingredient to maintaining enough precision to keep a pushdown
abstraction. The semantic spaces for the machine are modified thusly:

@@

\paragraph{PDCFA paper versus implementation:} since PDCFA is a
simpler analysis than CFA2, it enjoys a polynomial complexity in the
monovariant case. Its final ${\mathcal O}(n^6)$ form is a naive ``step
all states'' widening, which yields poor results as evident in the
above example. The common widening uses a work-list to only step
states that need stepping like the frontier-based widening we use ---
the fact that the store is shared amongst all states makes this a
semantic difference, and not just an implementation difference. Doing
this bumps the complexity up to ${\mathcal O}(n^9)$. The available
\emph{implementations} of PDCFA indeed do this and are themselves
${\mathcal O}(n^9)$. These bounds are (rough) absolute worst case that
is highly unlikely, if ever, to happen in practice. The work-list
abstraction restricts value flow to fewer spurious paths, often
leading to a faster analysis than a typical ${\mathcal O}(n^3)$ 0CFA
~\citep{ianjohnson:vardoulakis-lmcs11,
  dvanhorn:Earl2012Introspective}.

@@

CFA2 is the first published analysis of a higher-order programming
language that could properly match calls and returns. We will show
that it fits well into the same presentation we gave for
PDCFA. Vardoulakis and Shivers had a clear goal of harnessing the
extra information a pushdown model provides to produce a
high-precision analysis that works well in practice. This resulted in
more than just the call/return matching of the previous section, which
is why we are showing the two separately. There are two orthogonal
features of the semantics:
\begin{enumerate}
\item{stack allocation for some bindings}
\item{strong updates on stack frames for resolved non-determinism}
\end{enumerate}
The first of these is an addition to the stack-less context. There is a
conservative pre-analysis that checks locally whether a binding will
never escape, and classifies references as able to use the stack frame
or not. Their criteria for a binding never escaping is that it is
never referenced in a function that is not its binder. This can be
extended in a language with more linguistic features; see Kranz's
thesis about register-allocatable bindings in the Orbit Scheme
compiler \citep{ianjohnson:kranz:thesis:1988}. A stackable reference
is one that appears in the body of the binding function, by which we
mean not within the body of a nested function. They use the information
that a binding never escapes to not bother allocating it in the
heap. This has the advantage of not changing the heap, and thus leads
to less propagation. The addition of these stack frames makes the
analysis exponential in theory, though in practice they have been
observed to decrease running time in most cases.

@@

CFA2 uses a ``local semantics'' that is similar to our segmenting
continuations at function boundaries, but gets stuck when it gets to a
point where it would need to ``return.'' It instead appeals to an
external, imperative algorithm for summarization to sew the function
calls and returns together. What we show here is almost wholly the
same, only embodied still in terms of a machine semantics and thus
more easily reasoned about. The analogy to memoization as
summarization guide us in the construction of the analysis in the
coming section.

@@

The meta-functions the semantics uses to extend and consult the heap
and stack use implicit information from the pre-analysis we described
above:

@@

Sidestepping fake rebinding does not need to be restricted to
stackable references, but that is what CFA2 does. Indeed, as soon as
the non-determinism has been determined, we could extend the stack
frame so any subsequent reference means what it meant previously in
the function. Look-up would then always try the stack frame before
falling back on the heap.

@@

The meta-functions mentioned in the fixed semantics all deal with the
addition of $\maddr$ to contexts. If the context in an $\rt$ frame is
approximate, we must return to all the continuations known for all the
contexts it approximates ($\returns$). At return boundaries, the memo
table must add the result to all the represented contexts
($\memoize$). At capture time, we strip the store in the $\rt$ frame
if there is one, and replace it with an address.

@@

A new temporal logic that itself understands well-bracketing of call
and returns for stating and validating propositions, NWTL
~\citep{ianjohnson:alur:nwtl:2007}, applies to visibly pushdown
systems, which can precisely model programming languages that only have
well-bracketed call and returns. The propositional form of the logic
has decidable satisfiability checking, and control-flow queries are
shallow propositions to pose in the logic. However, the satisfiability
problem is \textsc{2Exptime}-hard and the algorithm is not given in
the on-the-fly form that a higher-order language would need to
effectively answer control-flow queries without already knowing the
answers. The logic itself is an exciting new frontier for expressing
program correctness properties, and a higher-order version would be a
welcome addition to the functional programmer's tool belt.

@@

There is also the question of clients of these analyses. Most
obviously we would want to know where we can implement first-class
control more efficiently. In particular, we would want to have an
escape analysis to find first-class continuations that don't need to
be heap-allocated, and single target abort operations that can be
turned into long jumps after a computed stack unraveling. On top of
optimizations, there are security analyses. Greater control of the
stack gives us the ability to drill deep into context-sensitive
security properties and make precise predictions such as, ``is it
possible to send image data on the network outside the context of the
user's permission?'' This way a user can use an image taking/sharing
application without the application surreptitiously taking and sending
images the user doesn't know about.
